\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data using R packages}}

### 2012-01-10 package/smutans/inst/doc/smutans.Rnw
Just an update.

%\VignetteIndexEntry{Analyzing Streptococcus mutans RNA-Seq data}
%\VignettePackage{smutans}

@@

\usepackage{Sweave}
\usepackage[letterpaper]{geometry}
\usepackage{hyperref,graphicx}
\usepackage{color}

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=4,height=4.5}
\newcommand{\Robject}[1]{\texttt{#1}}
\newcommand{\Rpackage}[1]{\textit{#1}}
\newcommand{\Rclass}[1]{\textit{#1}}
\newcommand{\Rfunction}[1]{{\small\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

@@

\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-Seq data using R packages}}}

@@

\begin{abstract}
We analyze a number of RNA samples from \textit{Streptococcus mutans} using R
packages: \Rpackage{DESeq}, \Rpackage{edgeR}, and \Rpackage{goseq}.
Tasks include the analysis of count data from RNA-Seq for
the detection of differentially expressed genes.  Tests of enrichment for
differentially expressed genes are performed using \Rpackage{goseq}.  The
package \Rpackage{smutans} contains the procedures.  This vignette explains the
use of the package. 
\end{abstract}

@@

The \Rpackage{smutans} package expects count data in the form of
\Rclass{CountDataSet}. Using BWA and samtools we aligned short reads from RNA
sequencing. Using gene annotations in the UA159 genome we counted short reads
that overlapped the genes.  The count values are raw counts of sequencing reads.
\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
reads that allows to assess the measurement precision correctly.  We did not
have rounded values of normalized counts, or counts of covered base pairs.

Several exeriments of \textit{S.\ mutans} RNA sampling were performed.
\fixme{Find the papers that are related with the experiments.} One of them was
presented in the object called \Robject{smutansGenes}.  The
\Robject{smutansGenes} object is of class \Rclass{CountDataSet}, which is the
data container used by \Rpackage{DESeq}. We have a covenient class called
\Rclass{Smutans}. The class contains slots for \Rclass{CountDataSet} objects,
and results from analyses.  Let's load the needed packages and the data as
follows.

We can create \Rclass{CountDataSet} objects using functions that are not
exported. One should create an R script at the directory that contains the
source code of the \Rpackage{smutans}. Put the following scripts in the R script
and run the script at the directory. That should create RData files at
the directory smutans/data: sm835Genes.RData, smutansGenes.RData.
<<samplePreparation,eval=FALSE>>=
#not run
library(DESeq)
source("smutans/R/core.R")
smutans.prepareData835NPP( )
smutans.prepareDataUA159TW1( ) 
smutans.prepareGoseq( )
q("no")
@

<<options,results=hide,echo=FALSE>>=
options(digits=3)
@

<<smutans2,results=hide>>=
library( "smutans" )
data( "smutansGenes" )
@

There are \Sexpr{length(colnames(counts(smutansGenes)))} RNA-seq samples in
\Robject{smutansGenes} that can be loaded from the package.
\Robject{smutansGenes} contains the counts and also metadata about the samples:

<<look_into_smutansGenes>>=
head( counts(smutansGenes) )
@

<<look_into_smutansGenes>>=
pData( smutansGenes )
@

The samples differ by experimental condition (glucose or
galactose) and by library type (ua159 or tw1). First, we only look at the ua159
data for now. We will look at the tw1 data later. 
We create an \Rclass{Smutans} object using the \Rclass{CountDataSet} object,
\Robject{smutansGenes}.

%--------------------------------------------------
\section{Differential expression analysis} \label{sec:de}
%--------------------------------------------------

%------------------------------------------------------------
\subsection{Inference: Using the UA159 samples for condition effect} 
\label{subsec:ua159only}
%------------------------------------------------------------

<<ua159onlySamples>>=
ua159only <- newSmutans( smutansGenes )
ua159only <- smutans.de2type( ua159only, type="ua159", 
                              condA="glucose", condB="galactose" )
@

\Rfunction{smutans.de2type} uses a two-factor data set to test differential
expression of genes. We compare 4 pairwise comparisons using a two-factor data
set. A pair of comparisons between glucose and galactose given either type ua159
or tw1. Another pairs of comparisons between types ua159 and tw1 given either
condition glucose or galactose. We focus on one of the four pairwise comparisons
first. Later, we will perform analyses for the rest of the three comparisons.
Note that the \Rclass{Smutans} contains \Rclass{CountDataSet} objects.
\Rfunction{smutans.de2type} estimates the effective library size, and the
dispersions using functions featured by \Rpackage{DESeq}. Then, it tests for
differential expression.

Because \Rpackage{DESeq} recommended to check
the dispersion estimates and to make sure that the data quality is as
expected, we plot the per-gene estimates against the normalized mean
expressions per gene, and then overlay the fitted curve in red
(Fig.\ \ref{figFit}).

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFit}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength.}
\label{figFit}
\end{figure}
<<figFit,fig=TRUE>>=
smutans.plotDispersionEstimates(ua159only)
@

As we estimated the dispersion from only two samples, we should expect the
estimates to scatter with quite some sampling variance around their true values.
Hence, \Rpackage{DESeq} should not use the per-gene estimates directly in the
test, because using too low dispersion values leads to false positives.  Many of
the values below the red line are likely to be underestimates of the true
dispersions, and hence, it is prudent to instead rather use the fitted value.
On the othe hand, not all of he values above the red line are overestimations,
and hence, the conservative choice is to keep them instead of replacing them
with their fitted values.

Having estimated the dispersion for each gene, we look for differentially
expressed genes. To contrast two conditions, e.g., to see whether there is
differential expression between conditions ``glucose'' and ``galactose'',
\Rfunction{smutans.de2type} already performed the function
\Rfunction{nbinomTest} provided by \Rpackage{DESeq}.  Let us first plot the
$\log_2$ fold changes against the base means, colouring in red those genes that
are significant at 10\% FDR.

For each gene, we get its mean expression level (at the base scale) as a joint
estimate from both conditions, and estimated separately for each  condition, the
fold change from the first to the second condition, the logarithm (to basis 2)
of the fold change, and the $p$ value for the statistical significance of this
change. The \texttt{padj} column contains the $p$ values, adjusted for multiple
testing with the Benjamini-Hochberg procedure (see the R function
\Rfunction{p.adjust}), which controls false discovery rate (FDR).

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figDE}
\caption{Plot of normalised mean versus $log_2$ fold change
(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
versus ``galactose''.}
\label{figDE}
\end{figure}
<<figDE,fig=TRUE>>=
smutans.plotDiffExp(ua159only)
@ %$
See Figure~\ref{figDE} for the plot.

We continue to analyze the data set using type tw1 to compare the two
conditions. Now, it should be straightfoward to perform the single-factor
analysis. Figures \ref{figFittw1only} and \ref{figDEtw1only} shows the
dispersion distribution and MA-plot for the contrast ``glucose'' versus
``galactose'' using type tw1 only. 
<<tw1onlySamples>>=
tw1only <- newSmutans( smutansGenes )
tw1only <- smutans.de2type( tw1only, type="tw1", 
                            condA="glucose", condB="galactose" )
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFittw1only}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength.}
\label{figFittw1only}
\end{figure}
<<figFittw1only,fig=TRUE>>=
smutans.plotDispersionEstimates(tw1only)
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figDEtw1only}
\caption{Plot of normalised mean versus $log_2$ fold change
(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
versus ``galactose''.}
\label{figDEtw1only}
\end{figure}
<<figDEtw1only,fig=TRUE>>=
smutans.plotDiffExp(tw1only)
@

Note that \Rfunction{smutans.de2} already performed the negative bionomial test
for each of the genes. We fetch adjusted p-values from the single-factor
analysis and the two-factor one to compare the number of genes that are found to
be differentially expressed using the same FDR rate of 0.1.
<<compareUA159onlyAndTW1only>>=
padj.ua159only <- smutans.padj( ua159only )
padj.tw1only <- smutans.padj( tw1only )
tabUA159onlyTW1only = table( "ua159 only" = padj.ua159only < .1, "tw1 only" = padj.tw1only < .1 )
addmargins( tabUA159onlyTW1only )
@
Of the \Sexpr{sum(tabUA159onlyTW1only)} gene, \Sexpr{tabUA159onlyTW1only[2,2]}
genes are found in both analyses. The analysis using all of tw1 only found
\Sexpr{tabUA159onlyTW1only[1,2]} genes. The analysis using only ua159 found
\Sexpr{tabUA159onlyTW1only[2,1]} genes.

Until now we focused on the effect of type not condition. 
<<glucoseOnlySamples>>=
glucoseOnly <- newSmutans( smutansGenes, title="glucoseOnly" )
glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
                            condA="ua159", condB="tw1" )
galactoseOnly <- newSmutans( smutansGenes, title="galactoseOnly" )
galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
                              condA="ua159", condB="tw1" )
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFitGlucoseOnly}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength. \textbf{figFitGlucoseOnly}}
\label{figFitGlucoseOnly}
\end{figure}
<<figFitGlucoseOnly,fig=TRUE>>=
smutans.plotDispersionEstimates(glucoseOnly)
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figDEGlucoseOnly}
\caption{Plot of normalised mean versus $log_2$ fold change
(this plot is sometimes also called the ``MA-plot'') for the contrast ``ua159''
versus ``tw1''. \textbf{figDEGlucoseOnly}}
\label{figDEGlucoseOnly}
\end{figure}
<<figDEGlucoseOnly,fig=TRUE>>=
smutans.plotDiffExp(glucoseOnly)
@

Note that \Rfunction{smutans.de2} already performed the negative bionomial test
for each of the genes. We fetch adjusted p-values from the single-factor
analysis and the two-factor one to compare the number of genes that are found to
be differentially expressed using the same FDR rate of 0.1.

\fixme{Two objects and one qvalue cutoff can be handled by a function. Let's
make a function in core.R.}
<<compareGlucoseOnlyAndGalactoseOnly>>=
#padj.glucoseOnly <- smutans.padj( glucoseOnly )
#padj.galactoseOnly <- smutans.padj( galactoseOnly )
#tabGlucoseOnlyGalactoseOnly = table( "glucose only" = padj.glucoseOnly < .1, 
#                                     "galactose only" = padj.galactoseOnly < .1 )
tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
                                                      galactoseOnly )
addmargins( tabGlucoseOnlyGalactoseOnly )
@
Of the \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} gene, \Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]}
genes are found in both analyses. The analysis using all of galactose only found
\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} genes. The analysis using glucose only found
\Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} genes.

%------------------------------------------------------------
\subsection{Inference: Using all of the samples for condition effect} 
\label{subsec:allSample}
%------------------------------------------------------------

% We must have at least one citation \cite{Anders:2010:GB}.

We could compare the two conditions, glucose and galactose, using either type
ua159 or tw1. This may or may not be desirable depending on the experimental
design. Using more samples might be more powerful in detecting differentially
expressed genes. Because we have a two-factor data set, \Robject{smutansGenes},
we wish to determine which genes are differentially expressed in one of the two
factors. Here, we first find genes differentially expressed in different
conditions of glucose and galactose using both types of ua159 and tw1. We will
compare those genes and those found in the single type analysis that we did
previously. Let's start the analysis by creating a new \Rclass{Smutans} object,
called \Robject{allSample}. 
<<allSamples>>=
allSample <- newSmutans( smutansGenes )
allSample <- smutans.de2( allSample )
@
\Rfunction{smutans.de2} is the basic function that uses a two-factor data set to
perform two-factor analysis. The two factors are type and condition. We are
interested in condition effect. We first check 
the dispersion estimates and to make sure that the data quality is as
expected, we plot the per-gene estimates against the normalized mean
expressions per gene, and then overlay the fitted curve in red
(Fig.\ \ref{figFit2}).

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFit2}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength.}
\label{figFit2}
\end{figure}

<<figFit2,fig=TRUE>>=
smutans.plotDispersionEstimates(allSample)
@

\fixme{Say something about Figure \ref{figFit2}.}

Note that \Rfunction{smutans.de2} already performed the negative bionomial test
for each of the genes. We fetch adjusted p-values from the single-factor
analysis and the two-factor one to compare the number of genes that are found to
be differentially expressed using the same FDR rate of 0.1.
<<compareUA159onlyAndAllSample>>=
padj.ua159only <- smutans.padj( ua159only )
padj.allSample <- smutans.padj( allSample )
tab = table( "ua159 only" = padj.ua159only < .1, "all samples" = padj.allSample < .1 )
addmargins( tab )
@
Of the \Sexpr{sum(tab)} gene, \Sexpr{tab[2,2]} genes are found in both analyses. The analysis using all
of the sample found \Sexpr{tab[1,2]} genes. The analysis using only ua159 found \Sexpr{tab[2,1]} genes. 
Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from the
two-factor analysis versus those from the single-factor analysis. The p-values
from the two-factor analysis are overall smaller than those from the
single-factor analysis. We would expect that this was the case because of higher
power of the two-factor analysis.

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figCompareUA159onlyAndAllSample}
\caption{Comparison of $p$ values (unadjusted) from the test using only the
four paired-end samples and the test using all seven samples. We can see that
the latter tend to be smaller, as expected from the higher power of the test with all samples.}
\label{figCompareUA159onlyAndAllSample}
\end{figure}

<<figCompareUA159onlyAndAllSample,fig=TRUE>>=
pval.ua159only <- smutans.pval( ua159only )
pval.allSample <- smutans.pval( allSample )
bottom = function(x, theta=1e-12) pmax(x, theta)
plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
abline(a=0, b=1, col="blue")
@

\fixme{Check the fitted regression using type and condition. The following are
from \Rpackage{DESeq} document.} 
The first three columns show the fitted coefficients, converted to a logarithm base 2 scale.
The log2 fold change due to the condition is shown in the third column. As indicated
by the column name, it is the effect of ``untreated'', i.e., the log ratio of
``untreated'' versus ``treated''. (This is unfortunately the other way round as
before, due to the peculiarities of contrast coding.) Note that the library type also
had noticeable influence on the expression, and hence would have increased the dispersion
estimates (and so reduced power) if we had not fitted an effect for it.
The column \emph{deviance}
is the deviance of the fit. (Comparing the deviances with a $\chi^2$ likelihood ratio test is
how \Rfunction{nbinomGLMTest} calculates the $p$ values.) The last column, \emph{converged},
indicates whether the calculation of coefficients and deviance has fully converged.
(If it is false too often, you can try to change the GLM control parameters, as
explained in the help page to \Rfunction{fitNbinomGLMs}.)

Finally, we show that taking the library type into account was important to have
good detection power by doing the analysis again using the standard workflow,
as outlined earlier, and without informing \Rpackage{DESeq} of the library types:

<<allSampleWithoutType>>=
allSampleWithoutType <- newSmutans( smutansGenes )
allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
                                                condA="glucose", 
                                                condB="galactose" )
@

<<compareUA159onlyAndAllSample>>=
padj.allSampleWithoutType <- smutans.padj( allSampleWithoutType )
tab2 <- table(
   `all samples simple` = padj.allSampleWithoutType < 0.1,
   `all samples GLM`    = padj.allSample < 0.1 )
addmargins( tab2 )
largeNumber <- 10
@
Of the \Sexpr{sum(tab2)} gene, \Sexpr{tab2[2,2]} genes are found in both analyses. The analysis using all
of the sample found \Sexpr{tab2[1,2]} genes. The analysis using all of the same
sample but without considering type found \Sexpr{tab2[2,1]} genes.

>>>>

%\VignetteIndexEntry{Analyzing Streptococcus mutans RNA-seq data}
%\VignettePackage{smutans}

@@

\usepackage{Sweave}
\usepackage[letterpaper]{geometry}
\usepackage{hyperref,graphicx}
\usepackage{color}
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=4,height=4.5}
\newcommand{\Robject}[1]{object \texttt{#1}}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\Rclass}[1]{class \textit{#1}}
\newcommand{\Rfunction}[1]{{function \small\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

@@

\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data using R packages}}}

@@

\begin{abstract}
We analyze a number of RNA samples from \textit{Streptococcus mutans}.

using 
\Rpackage{DESeq}, \Rpackage{edgeR}, and \Rpackage{goseq}.
Tasks include the analysis of count data from RNA-seq for
detection of differentially expressed genes, and tests of enrichment for
differentially expressed genes using \Rpackage{goseq}.

We use RNA-seq data to define genes whose expression changes between
samples \cite{Oshlack2010}. Short reads are first mapped to a
reference genome using approaches such as BWA \cite{Li2009a}.
Reads fitting into one of the annotated intervals are counted.
These counts are passed to a statistical package which treats differences
in read counts for statistical significance. For this purpose, we made
use of \Rpackage{DEseq} \cite{Anders2010}. This prodecure returns a list of
differentially expressed genes, which can be passed to downstream
analysis, such as functional category enrichment test using 
\Rpackage{goseq} \cite{Young2010}.

This
\Rpackage{smutans} contains procedures that make use of functions
provided by the above R packages.  
\end{abstract}

@@

The \Rpackage{smutans} expects count data in the form of
\Rclass{CountDataSet}, which is a type provided by \Rpackage{DESeq}.  Using BWA
\cite{Li2009a} and samtools \cite{Li2009b} 
we aligned short reads from RNA sequencing. Using gene annotations
in the UA159 genome we counted short reads that overlapped the genes.  The count
values are raw counts of sequencing reads. The \Rpackage{DESeq}'s statistical
model, or negative binomial model, hold for raw reads that allows to assess the
measurement precision correctly; we did not have rounded values of normalized
counts, or counts of covered base pairs.  Several exeriments of \textit{S.\
mutans} RNA sampling were performed.  \fixme{Find the papers that are related
with the experiments.} One of them was presented in the 
\Robject{smutansGenes} as a part of \Rpackage{smutans}. The
\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the
data type used by \Rpackage{DESeq}. We have a convenient 
\Rclass{Smutans}, which contains slots for \Rclass{CountDataSet} objects, and
results from analyses thereof.  Before loading necessary R packages and data
sets, let us describe what functions were used for preparing data objects such as
\Robject{smutansGenes}. One should create an R script file at the directory that contains the
source code directory of the \Rpackage{smutans}. Put the following scripts in the R script
file
and run the script at that directory. That should create RData files at
the directory smutans/data. User should not run the R script because the data
files should be found in the \Rpackage{smutans}.
% http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
% http://www.oralgen.lanl.gov/cgi-bin/gene_id_search.cgi?dbname=smut&gene_id=SMu0001
% \fixme{Where are flat files for genes of S. mutans UA159?}
<<samplePreparation,eval=FALSE>>=
# not run
# but create a R scirpt file with the following scripts, and
# run it, i.e., Rscript run.R
library(DESeq)
source("smutans/R/core.R")
smutans.prepareData34( )
smutans.prepareData835NPP( )
smutans.prepareDataUA159TW1( ) 
smutans.prepareGoseq( )
smutans.prepareDataOMZ175( ) 
smutans.prepareDataUA159CSP( ) 
smutans.prepareDataSMU86CSP( )
q("no")
@
<<options,results=hide,echo=FALSE>>=
options(digits=3)
@
Now, see some of the data objects. Load the \Rpackage{smutans}, and a data
\Robject{smutansGenes}. Note that the object is of type \Rclass{CountDataSet} of
\Rpackage{DESeq} in which users should find \Rfunction{counts} and
\Rfunction{pData}. 
<<smutans2,results=hide>>=
library( "smutans" )
data( "smutansGenes" )
@
There are 
\Sexpr{length(colnames(counts(smutansGenes)))}
RNA-seq samples in \Robject{smutansGenes}, which
contains the counts and also metadata about the samples:
<<look_into_smutansGenes>>=
head( counts(smutansGenes) )
@
<<look_into_smutansGenes>>=
pData( smutansGenes )
@
The samples differ by experimental condition (``glucose'' or
``galactose'') and by library type (``ua159'' or ``tw1'').

Samples of 835NP, 835P, and UA159 no CSP should be loaded using the following
command.
<<sm835,results=hide>>=
data( "sm835Genes" )
@

\fixme{Describe other data objects as well.
sm34Genes.RData,
sm835Genes.RData,
smutans.cat.desc.RData,
smutans.feature.genes.RData, and
smutans.go.genes.RData.}

%--------------------------------------------------
\section{DE analysis using UA159 and TW1 samples with
glucose and galactose} \label{sec:detw1}
%--------------------------------------------------

We could compare four pairwise comparisons using the two-factor data set of
\Robject{smutansGenes}: a pair of comparisons between ``glucose'' and
``galactose'' given either type ``ua159'' or ``tw1'', and another pair of
comparisons between types ``ua159'' and ``tw1'' given either condition
``glucose'' or ``galactose.'' We focus on one of the four pairwise comparisons
first. Later, we will perform analyses for the rest of the three comparisons.
We look at the ``ua159'' data for now, and will look at the ``tw1'' data later.

We can cluster samples using variance stabilized data. Figure \ref{figClust}
shows the \Sexpr{length(colnames(counts(smutansGenes)))} sampled clustered. 
The clustering should reflect our experimental design, i.e., samples are
more similar when they have the same treatment or the same library type.
To make this conclusion, we estimate the dispersion with
mode ``blind'' (see \Rpackage{DESeq} vignette), as only then,
the variance stabilizing transformation is not informed about the design and
hence not biased towards a result supporting it. Such an analysis is useful
for quality control.  This appears to be useful as a first in any analysis.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{smutans-figClust}
\caption{Heatmap showing the Euclidean distances between the samples
  as calculated from the variance-stabilizing transformation of the
  count data.}
\label{figClust}
\end{figure}
<<figClust,fig=TRUE>>=
allSample <- newSmutans( smutansGenes )
smutans.de2Clust( allSample )
@
As expected, two experimental conditions of ``glucose'' and ``galactose'' divide the
samples. Within each condition, two different types of ``ua159'' and ``tw1'' divide
them again. The pair of ``glucose ua159'' and ``galactose tw1'' is the most
distinguished.

%------------------------------------------------------------
\subsection{Inference using single-factor data} 
\label{subsec:ua159only}
%------------------------------------------------------------

We create a \Rclass{Smutans} object using the \Rclass{CountDataSet}
\Robject{smutansGenes}. The title option would be useful in distinguishing
different objects of \Rclass{Smutans} as we have more R objects with the type.  
A subsequent call to
\Rfunction{smutans.de2} uses a two-factor data set to test differential
expression of genes.  The \Rfunction{smutans.de2} estimates effective library
sizes, and dispersions using functions provided by \Rpackage{DESeq}. Then, it
tests genes for differential expression.
<<ua159onlySamples>>=
ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
ua159only <- smutans.de2( ua159only, type="ua159", 
                          condA="glucose", condB="galactose" )
@
Because \Rpackage{DESeq} recommended to check
the dispersion estimates and to make sure that the data quality is as
expected, we plot the per-gene estimates against the normalized mean
expressions per gene, and then overlay the fitted curve in red color
(Fig.\ \ref{figFit}).

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFit}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength for the data set of UA159-only sample
to compare glucose and galactose conditions.}
\label{figFit}
\end{figure}
<<figFit,fig=TRUE>>=
smutans.plotDispersionEstimates(ua159only)
@
\fixme{The following paragraph is adopted from \Rpackage{DESeq} vignette.}
As we estimated the dispersion from only two samples, we should expect the
estimates to scatter with quite some sampling variance around their true values.
Hence, \Rpackage{DESeq} should not use the per-gene estimates directly in the
test, because using too low dispersion values leads to false positives.  Many of
the values below the red line are likely to be underestimates of the true
dispersions, and hence, it is prudent to instead rather use the fitted value.
On the othe hand, not all of the values above the red line are overestimations,
and hence, the conservative choice is to keep them instead of replacing them
with their fitted values.
Having estimated the dispersion for each gene, we look for differentially
expressed genes. To contrast two conditions, e.g., to see whether there is
differential expression between conditions ``glucose'' and ``galactose'',
\Rfunction{smutans.de2} already performed the 
\Rfunction{nbinomTest} provided by \Rpackage{DESeq}.  
Let's list genes with resulting information from \Rpackage{DESeq}.
<<ua159onlyDE2List>>=
head(smutans.de2List( ua159only ))
@
For each gene, we get its mean expression level (at the base scale) as a joint
estimate from both conditions, and estimated separately for each condition, the
fold change from the first to the second condition, i.e., 
$\frac{baseMeanB}{baseMeanA}$, the logarithm (to basis 2)
of the fold change, and the $p$ value for the statistical significance of this
change. The \texttt{padj} column contains the $p$ values, adjusted for multiple
testing with the Benjamini-Hochberg procedure (see the R function
\Rfunction{p.adjust}), which controls false discovery rate (FDR). Later, we
print the gene list out in a CSV file using option file of
\Rfunction{smutans.de2List}.
%-----------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figDE}
\caption{Plot of normalised mean versus $log_2$ fold change
(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
versus ``galactose'' using UA159-only sample.}
\label{figDE}
\end{figure}
%-----------------------------------------------------------------------------
<<figDE,fig=TRUE>>=
smutans.plotDiffExp(ua159only)
@
Figure \ref{figDE} shows a plot of the $\log_2$ fold changes against the base
means, coloring in red those genes that are significant at 10\% of FDR. 
Let us use ``tw1'' type to compare ``glucose'' and ``galactose'' conditions.
We continue to analyze the data set using type ``tw1'' to compare the two
conditions. We perform the single-factor analysis. 
<<tw1onlySamples>>=
tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
tw1only <- smutans.de2( tw1only, type="tw1", 
                        condA="glucose", condB="galactose" )
tabUA159onlyTW1only <- smutans.compareResult( ua159only, tw1only )
addmargins( tabUA159onlyTW1only )
@
The table, \Robject{tabUA159onlyTW1only}, shows numbers of genes found to be 
differentially expressed in the two analyses: ``UA159 Only'' and ``TW1 Only.''
Of the total \Sexpr{sum(tabUA159onlyTW1only)} genes, \Sexpr{tabUA159onlyTW1only[2,2]}
genes were found in both analyses. The analysis using TW1-only sample found
\Sexpr{tabUA159onlyTW1only[1,2]} genes. The analysis using UA159-only sample found
\Sexpr{tabUA159onlyTW1only[2,1]} genes.

Until now we focused on the effect of condition either using UA159-only or
TW1-only samples. We now change the gear to see effects of sample types on
differentially expressed genes. We create two \Rclass{Smutans} objects:
\Robject{glucoseOnly} and \Robject{galactoseOnly}, and compare them in detecting
differentially expressed genes:
<<glucoseOnlySamples>>=
glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
                            condA="ua159", condB="tw1" )
galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
                              condA="ua159", condB="tw1" )
tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
                                                      galactoseOnly )
addmargins( tabGlucoseOnlyGalactoseOnly )
@
Of the total \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} genes,
\Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]} genes were found in both analyses at FDR
rate of 10\%. The analysis using galactose-only sample found
\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} genes. The analysis using glucose-only
sample found \Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} genes. 
Table \ref{tab:de} summarizes the number of differentially expressed genes that
were found using single-factor analyses at FDR of 10\%.

%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Numbers of differentially expressed genes in different analyses}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{ccccl}
\hline 
Sample A & Sample B & Number of Genes & Result File \\
\hline 
UA159 Glucose & UA159 Galactose & \Sexpr{sum(tabUA159onlyTW1only[2,])} & ua159only.csv \\
TW1 Glucose & TW1 Galactose & \Sexpr{sum(tabUA159onlyTW1only[,2])} & tw1only.csv \\
UA159 Glucose & TW1 Glucose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[2,])} & glucoseOnly.csv \\
UA159 Galactose & TW1 Galactose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[,2])} & galactoseOnly.csv \\
\hline 
\end{tabular}
\par\end{centering}
\label{tab:de}
\end{table}
%-----------------------------------------------------------------------------

%------------------------------------------------------------
\subsection{Inference using all of the samples} 
\label{subsec:allSample}
%------------------------------------------------------------

It is inconvenient to use four different analyses to find differentially
expressed genes under one of two factors. Besides, it would be desirable to use all of
the data to find differentially expressed genes because using more samples might
be more powerful in detecting differentially expressed genes.  Let us find genes
differentially expressed in different conditions of ``glucose'' and
``galactose'' using both types of ``ua159'' and ``tw1.'' We will compare these
genes and those found in the single-factor analysis that we did previously,
i.e., \Robject{ua159only}.
Let's start with multi-factor the analysis by creating a new \Rclass{Smutans} object, called
\Robject{allSample}. Although we already used \Robject{allSample} when
clustering the samples, we recreate it for instructive purposes.
<<allSamples,results=hide>>=
allSample <- newSmutans( smutansGenes, title="All Samples for Comparing Conditions" )
allSample <- smutans.de2( allSample )
@
The \Rfunction{smutans.de2} uses a two-factor data set to
perform two-factor analysis. The two factors are type and condition. We are here
interested in condition effect. We could check 
the dispersion estimates and to make sure that the data quality is as
expected by plotting the per-gene estimates against the normalized mean
expressions per gene, and then overlay the fitted curve.
<<figFitAllSample,eval=FALSE>>=
# not run
smutans.plotDispersionEstimates(allSample)
@
%-----------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figCompareUA159onlyAndAllSample}
\caption{Comparison of $p$ values (unadjusted) from the test using only the
four paired-end samples and the test using all seven samples. We can see that
the latter tend to be smaller, as expected from the higher power of the test with all samples.}
\label{figCompareUA159onlyAndAllSample}
\end{figure}
%-----------------------------------------------------------------------------
<<figCompareUA159onlyAndAllSample,fig=TRUE>>=
pval.ua159only <- smutans.pval( ua159only )
pval.allSample <- smutans.pval( allSample )
bottom = function(x, theta=1e-12) pmax(x, theta)
plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
abline(a=0, b=1, col="blue")
@
<<compareUA159onlyAndAllSample>>=
tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
                                                   allSample )
addmargins( tabUA159onlyAndAllSample )
@
Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
genes.  Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from
the two-factor analysis versus those from the single-factor analysis. The
p-values from the two-factor analysis are overall smaller than those from the
single-factor analysis. We would expect that this was the case because of higher
power of the two-factor analysis.

We can show that taking the library type into account was important to have
good detection power by doing the analysis again using the standard workflow,
as outlined earlier, and without informing \Rpackage{DESeq} of the library
types. This anlaysis is not necessary. However, it clearly shows the advantage of the
two-factor analysis.
<<allSampleWithoutType>>=
allSampleWithoutType <- newSmutans( smutansGenes, "All Sample without Type" )
allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
                                                condA="glucose", 
                                                condB="galactose" )
tabWithoutWithAllSample <- smutans.compareResult( allSampleWithoutType, 
                                                  allSample )
addmargins( tabWithoutWithAllSample )
@
Of the total \Sexpr{sum(tabWithoutWithAllSample)} genes,
\Sexpr{tabWithoutWithAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabWithoutWithAllSample[1,2]}
genes. The analysis using all of the same sample but without considering type
found only \Sexpr{tabWithoutWithAllSample[2,1]} genes.
